FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2018-05-28 ECOOP18/sections/Formalization.tex
revision

\subsubsection{Interfaces}
In order to achieve multiple inheritance, an interface can have a set of 
parent interfaces, where such a set can be empty. The interface declaration $\interface{I}{I}{M}$ introduces an interface named $I$ with parent interfaces $\overline{I}$ and a suite of methods $\overline{M}$. The methods of $I$ may either override methods that are already defined in $\overline{I}$ or add new functionality special to $I$, we will illustrate this in more detail later.

>>>>

\subsubsection{Interfaces}
\yanlin{Bruno: acyclic, clarified herer.}
In order to achieve multiple inheritance, an interface can have a set of 
parent interfaces, where such a set can be empty. Moreover, as usual
in class-based languages, the class hierarchy is acyclic. The interface declaration $\interface{I}{I}{M}$ introduces an interface named $I$ with parent interfaces $\overline{I}$ and a suite of methods $\overline{M}$. The methods of $I$ may either override methods that are already defined in $\overline{I}$ or add new functionality special to $I$, we will illustrate this in more detail later.

### 2018-05-28 ECOOP18/sections/Introduction.tex
revision

To present hierarchical overriding and dispatching, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. 
In summary, our contributions are:
\begin{itemize}
	\item \textbf{A formalization of the hierarchical dispatching algorithm} that integrates both the static type and dynamic type for method dispatch, and 
	ensures unambiguity as well as extensibility in the presence
        of unintentional method conflicts.
	\item \textbf{Hierarchical overriding:} a novel notion that allows
          methods to override individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on
          Featherweight Java, supporting the above features. 
          We provide the static and dynamic semantics and prove the
          type soundness of the model.
	\item \textbf{Prototype implementation\footnote{The implementation is available at \url{https://github.com/YanlinWang/MIM/tree/master/Calculus}}:} a
          simple implementation of \MIM{} interpreter in Scala. The implementation can type-check and run variants
          of all the examples shown in this paper.
\end{itemize}

>>>>

To present hierarchical overriding and dispatching, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. We also have a prototype implementation of
a \name{} interpreter written in Scala. The implementation validates
all the examples presented in the paper. 
One nice feature of the implementation is that it can show the
detailed step-by-step evaluation of the program, which is convenient
for understanding and debugging programs \& semantics. 
%The final model of \name{}
%is actually a bit different from our initial models. This is partly because in
%the process of implementing and proving theorems we found errors and modified the earlier models, creating a positive feedback circle.

In summary, our contributions are:
\begin{itemize}
	\item \textbf{A formalization of the hierarchical dispatching algorithm} that integrates both the static type and dynamic type for method dispatch, and 
	ensures unambiguity as well as extensibility in the presence
        of unintentional method conflicts.
	\item \textbf{Hierarchical overriding:} a novel notion that allows
          methods to override individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on
          Featherweight Java, supporting the above features. 
          We provide the static and dynamic semantics and prove the
          type soundness of the model.
	\item \textbf{Prototype implementation\footnote{The implementation is available at \url{https://github.com/YanlinWang/MIM/tree/master/Calculus}}:} a
          simple implementation of \MIM{} interpreter in Scala. The implementation can type-check and run variants
          of all the examples shown in this paper.
\end{itemize}

### 2018-05-28 ECOOP18/sections/Overview.tex
revision

In the rest of the paper, we use a Java-like syntax for programs. All types are defined with the keyword
\lstinline|interface|; the concept is closely related to Java 8
interfaces with default methods~\cite{bono14} and traits. In short, 
an interface in our model has the following characteristics:
\begin{itemize}
	\item It allows multiple inheritance.
	\item Every method is either abstract or implemented with a body (like Java 8 default methods). 
	\item The \lstinline|new| keyword is used to instantiate an interface.
	\item It cannot have state. But it is possible to support state in future work.
\end{itemize}
%In the remainder of this section we show three problems that arize 
%from unintentional method conflicts. The first two problems

@@

With unintentional method conflicts, however, the situation is more
complicated because different, separate, conflicting methods can coexist
in one class. Ideally, we would like to support overriding for those
methods too, in exactly the same way that overriding is available for 
other (non-conflicting) methods. However, we need to be able to
override the individual conflicting methods, rather than overriding all
conflicting methods into a single merged one. We illustrate the
problem and the need for a more refined overriding mechanism with 
an example.
Suppose that the programmer defines a new interface \lstinline|DrawableSafeDeck| (based on the code in Section~\ref{subsec:problem2} without the old \lstinline|DrawableSafeDeck|), but he needs to override
\lstinline|Drawable.draw| and give a new implementation of drawing so that the deck can indeed be visualized on the canvas.

@@

\subsubsection{Potential solutions/workarounds in existing languages} 
\yanlin{Bruno: one reviewer says S2.3.1 is "less clear", could you check again?}Unfortunately in all languages we know of 
(including C++), the existing approaches are unsatisfactory.
One direction is to simply avoid this issue, by putting overriding before inheritance. For example, as shown in Figure~\ref{fig:drawablesafedeck}(middle), we define a new component \lstinline|DrawableRect| that extends \lstinline|Drawable|, which simply draws the deck as a rectangle, and modifies the hierarchy:
\begin{lstlisting}
interface DrawableRect extends Drawable {
  void draw() {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
interface DrawableSafeDeck extends DrawableRect, SafeDeck {}
\end{lstlisting}
This workaround seems to work, but there are severe issues:
\begin{itemize}
	\item It changes the hierarchy and existing code, hence breaks the modularity.
	\item Separate overriding is required to come after the \wordfork{} 
          inheritance, especially when the implementation needs functionality from both parents. In the above code, we have assumed
	that the overriding is unrelated to \lstinline|Deck|. But when the drawing relies on some information of the \lstinline|Deck|
	object, we have to either introduce field methods for delegation or change the signature of \lstinline|draw| to take a parameter.
	Either way introduces unnecessary complexity and affects extensibility.
\end{itemize}

@@

The UML graph is shown in Figure~\ref{fig:drawablesafedeck}(right), where
the up-arrow $\uparrow$ is short for ``\lstinline|override|''.
Here the idea is that \emph{only} \lstinline{Drawable.draw} is
overridden. This is accomplished by specifying, in the method
definition, that the method only overrides the \lstinline|draw| 
from \lstinline|Drawable|.  
The individual overriding allows us to make use of the methods from \lstinline|SafeDeck| as well. In the formalization, the 
hierarchical overriding feature is an important feature, involved in the algorithm of hierarchical dispatch.
%\bruno{the last two sentences are a bit unclear. I think it is related to ``interaction''.}

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for unintentional conflicts, namely the \wordfork{} inheritance. (b) uses overriding to explicitly merge the conflicting methods. (c) represents hierarchical overriding. (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super method invocation} in a method body. The invocation \lstinline|super.T::m()| will ignore all the subtypes of \lstinline|T|, and only look at \lstinline|T| together with its super interfaces. It should behave the same as \lstinline|new T().m()| in principle.

>>>>

In the rest of the paper, we use a Java-like syntax for programs. All types are defined with the keyword
\lstinline|interface|; the concept is closely related to Java 8
interfaces with default methods~\cite{bono14} and traits. In short, 
an interface in our model has the following characteristics:
\begin{itemize}
	\item It allows multiple inheritance.
	\item Every method is either abstract or implemented with a body (like Java 8 default methods). 
	\item The \lstinline|new| keyword is used to instantiate an interface.
	\item It cannot have state. 
\end{itemize}
%In the remainder of this section we show three problems that arize 
%from unintentional method conflicts. The first two problems

@@

With unintentional method conflicts, however, the situation is more
complicated because different, separate, conflicting methods can coexist
in one class. Ideally, we would like to support overriding for those
methods too, in exactly the same way that overriding is available for 
other (non-conflicting) methods. However, we need to be able to
override the individual conflicting methods, rather than overriding all
conflicting methods into a single merged one.

We illustrate the
problem and the need for a more refined overriding mechanism with 
an example.
Suppose that the programmer defines a new interface \lstinline|DrawableSafeDeck| (based on the code in Section~\ref{subsec:problem2} without the old \lstinline|DrawableSafeDeck|), but he needs to override
\lstinline|Drawable.draw| and give a new implementation of drawing so that the deck can indeed be visualized on the canvas.

@@

\subsubsection{Potential solutions/workarounds in existing languages} 
\yanlin{Bruno: one reviewer says S2.3.1 is "less clear", could you
  check again?}\bruno{seems ok to me.}
Unfortunately in all languages we know of 
(including C++), the existing approaches are unsatisfactory.
One direction is to simply avoid this issue, by putting overriding before inheritance. For example, as shown in Figure~\ref{fig:drawablesafedeck}(middle), we define a new component \lstinline|DrawableRect| that extends \lstinline|Drawable|, which simply draws the deck as a rectangle, and modifies the hierarchy:
\begin{lstlisting}
interface DrawableRect extends Drawable {
  void draw() {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
interface DrawableSafeDeck extends DrawableRect, SafeDeck {}
\end{lstlisting}
This workaround seems to work, but there are severe issues:
\begin{itemize}
	\item It changes the hierarchy and existing code, hence breaks the modularity.
	\item Separate overriding is required to come after the \wordfork{} 
          inheritance, especially when the implementation needs functionality from both parents. In the above code, we have assumed
	that the overriding is unrelated to \lstinline|Deck|. But when the drawing relies on some information of the \lstinline|Deck|
	object, we have to either introduce field methods for delegation or change the signature of \lstinline|draw| to take a parameter.
	Either way introduces unnecessary complexity and affects extensibility.
\end{itemize}

@@

The UML graph is shown in Figure~\ref{fig:drawablesafedeck}(right), where
the up-arrow $\uparrow$ is short for ``\lstinline|override|''.
Here the idea is that \emph{only} \lstinline{Drawable.draw} is
overridden. This is accomplished by specifying, in the method
definition, that the method only overrides the \lstinline|draw| 
from \lstinline|Drawable|.  
The individual overriding allows us to make use of the methods from \lstinline|SafeDeck| as well. In the formalization, the 
hierarchical overriding feature is an important feature, involved in
the algorithm of hierarchical dispatch.

Note that, although the example here only shows one conflicting method
being overriden, hierarchical overriding allows (as expected) multiple conflicting
methods to be overriden in the same class.

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for unintentional conflicts, namely the \wordfork{} inheritance. (b) uses overriding to explicitly merge the conflicting methods. (c) represents hierarchical overriding. (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super
% method invocation} in a method body. The invocation
% \lstinline|super.T::m()| will ignore all the subtypes of
% \lstinline|T|, and only look at \lstinline|T| together with its
% super interfaces. It should behave the same as \lstinline|new
% T().m()| in principle.

\bruno{I remember that a big complain in the reviews was talking about
diamond inheritance. The Figure does have diamond inheritance but this
wasn't clear to the reviewers. Please expand the text to emphasize
diamond inheritance better!}
